Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2021-03-25 bbbq_article.tex

For MHC-I, 9-mers were used, as this is the length used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used, 
as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (Mycobacterium tuberculosis (MTb), UniProt ID UP000001584) 
reference proteome. The pathogens were chosen based 
on their impact on the human 
population \richel{ref}.
We used TMHMM  \richel{ref} to predict the topology 
of the proteins within the proteome.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; \richel{ref} for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
\richel{This is Frans' text: 
Unlike MHC-I where only one variable chain determines the landscape of the epitope binding pocket allowing for the classification of 5 HLA-A and 8 HLA-B superhaplotypes with .... variation in peptide binding repertoire, the MHC-II peptide binding groove is determined by both the alpha and beta chain. For HLA DQ both the alpha and beta chain is highly polymorphic  ..... \frans{why is HLA-DP not predicted?}
We took the alleles that represent percentage of the occurrence int he overall human population according to ... et al. 
}

@@

@@

Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.

[RJCB: How many SNPs in TMHs?]
See figure \ref{fig:f_tmh_ncbi} to see which percentages
are TMH of the transmembrane proteins.

@@

% \paragraph{MHC-II is expected to present TMHs}
\frans{we moeten iets schrijven over de verschillende haplotypen wat is the occurence van de haplotypen met andere worden hoe relevnet zijn de haplotypen die veel TMH presenteren?}

The over-presentation of TMHs for MHC-II haplotypes 
is more pronounced than for MHC-I,
as with a third of the predicted binders 
\richel{Hier ben ik nu met verwerken...}
from the human and MtB proteome stem from TMH human reference proteome and MtB reference proteome to be presented on MHC-II. \frans{ik zoek nog naar associaties} 
such as \verb;HLA-DQA1*0101/DQB1*0501;

@@

To determine the additional over-presentation of TMH-derived epitopes 
in MHC-II (as compared to MHC-I), we used the same normalized percentages
for the human proteome only and compared the averages between MHC-I and MHC-II.
See figure \ref{fig:fig_rel_presentation}.

>>>>

For MHC-I, 9-mers were used, as this is the length used in our earlier
study (\cite{bianchi2017}). For MHC-II, 14-mers were used, 
as these are the most frequently occurring
epitope length (\cite{bergseng2015different}).
We used a human (UniProt ID UP000005640\_9606), 
viral (SARS-CoV-2, UniProt ID UP000464024) 
and bacterial (Mycobacterium tuberculosis (MTb), UniProt ID UP000001584) 
reference proteome. The pathogens were chosen based 
on their impact on the human 
population \richel{ref}.
We used TMHMM  \richel{ref} to predict the topology 
of the proteins within the proteome.
To predict the affinity of an epitope to a certain MHC molecule,
we used \verb;EpitopePrediction; (\cite{bianchi2017}) for MHC-I 
and \verb;MHCnuggets; \richel{ref} for MHC-II.
The 13 MHC-I haplotypes used in this study are the same as 
used in our previous study (\cite{bianchi2017}).
The MHC-II haplotypes used additionally were selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional}),
resulting in 21 haplotypes.
\richel{This is Frans' text, merge it with the current text: 
Unlike MHC-I where only one variable chain determines the landscape of the epitope binding pocket allowing for the classification of 5 HLA-A and 8 HLA-B superhaplotypes with .... variation in peptide binding repertoire, the MHC-II peptide binding groove is determined by both the alpha and beta chain. For HLA DQ both the alpha and beta chain is highly polymorphic  ..... \frans{why is HLA-DP not predicted?}
We took the alleles that represent percentage of the occurrence int he overall human population according to ... et al. 
}

@@

HIERO
Additionally, we investigated if MHC-I and MHC-II are equally prone
to present TMH-derived epitopes. 
See \ref{fig:rel_presentation}
\ref{fig:rel_presentation_per_haplotype.png}

fig_rel_presentation_per_haplotype.png

@@

Using a alpha value of $0.05$, as is common practice for an uninformed
expectation, we conclude that there are significantly less SNPs
in TMHs as expected by chance.
See figure \ref{fig:f_tmh_ncbi} to see which percentages
are TMH of the transmembrane proteins.

@@

% \paragraph{MHC-II is expected to present TMHs}
\frans{we moeten iets schrijven over de verschillende haplotypen wat is the occurence van de haplotypen met andere worden hoe relevant zijn de haplotypen die veel TMH presenteren?}
The over-presentation of TMHs for MHC-II haplotypes 
is slight more pronounced than for MHC-I,
as with a third of the predicted binders 
\richel{Hier ben ik nu met verwerken...}
from the human and MtB proteome stem from TMH human reference proteome and MtB reference proteome to be presented on MHC-II. \frans{ik zoek nog naar associaties} 
such as \verb;HLA-DQA1*0101/DQB1*0501;

@@

To determine the additional over-presentation of TMH-derived epitopes 
in MHC-II (as compared to MHC-I), we used the same normalized percentages
for the human proteome only and compared the averages between MHC-I and MHC-II.
Figure \ref{fig:fig_rel_presentation_per_haplotype} shows the normalized
TMH-derived epitope presentation per haplotype.

See figure \ref{fig:fig_rel_presentation}.

\begin{figure}[!htbp]
  \includegraphics[width=\textwidth]{bbbq_1_smart_results/fig_rel_presentation_per_haplotype.png}
  \caption{
    Normalized proportion of MHC-I and MHC-II epitopes overlapping with TMHs
    for human proteomes.
  }
  \label{fig:fig_rel_presentation_per_haplotype}
\end{figure}
